---
title: "Lab week 2 - Stacking"
subtitle: "Data Science and Machine Learning 3 - CEU 2020"
author: "Jeno Pal"
date: '2019-02-26'
output:
  html_document:
    df_print: paged
  html_notebook:
    df_print: paged
---

```{r}
library(data.table)
library(GGally)

library(h2o)
h2o.init(max_mem_size='8g')
```


Stacking:

* specify different $L$ predictive models
* score training observations using out-of-fold predictions
* this gives a "level-one" data: the original outcome and the scores
coming from the $L$ base models
* use these to estimate a second level predictive model ("meta-learner")

When predicting new observations

  * the base models are estimated using all training observations, use these
  to get scores for the new observation
  * input this to the meta-learner model to get the predicted outcome

The more uncorrelated predictions are, the more room there is to
improve individual models.

Stacked models do not always perform better than individual ones but
many times they do.

It is very convenient to create stacked ensembles in h2o, we are
gaining practice with this now. See more [here](http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/stacked-ensembles.html).


```{r}
data <- h2o.importFile("../../data/airlines/airline100K.csv")

data_split <- h2o.splitFrame(data, ratios = 0.2, seed = 123)
data_train <- data_split[[1]]
data_test <- data_split[[2]]

y <- "dep_delayed_15min"
X <- setdiff(names(data_train), y)
```

Train some base learners using cross validation.

```{r}
glm_model <- h2o.glm(
  X, y,
  training_frame = data_train,
  family = "binomial",
  alpha = 1, 
  lambda_search = TRUE,
  seed = 123,
  nfolds = 5, 
  keep_cross_validation_predictions = TRUE  # this is necessary to perform later stacking
)
```

```{r}
gbm_model <- h2o.gbm(
  X, y,
  training_frame = data_train,
  ntrees = 200, 
  max_depth = 10, 
  learn_rate = 0.1, 
  seed = 123,
  nfolds = 5, 
  keep_cross_validation_predictions = TRUE
)
```
```{r}
deeplearning_model <- h2o.deeplearning(
  X, y,
  training_frame = data_train,
  hidden = c(32, 8),
  seed = 123,
  nfolds = 5, 
  keep_cross_validation_predictions = TRUE
)
```

```{r}
ensemble_model <- h2o.stackedEnsemble(
  X, y,
  training_frame = data_train,
  base_models = list(glm_model, 
                     gbm_model,
                     deeplearning_model),
  keep_levelone_frame = TRUE)
```

```{r}
# we can look into the level-1 features that are used as features by the meta learner
level_1_features <- h2o.getFrame(ensemble_model@model$levelone_frame_id$name)
level_1_features
```

```{r}
# inspect correlations among scores from different models
level_1_dt <- as.data.table(level_1_features)[, 1:3]
setnames(level_1_dt, c("glm", "gbm", "dl"))

ggcorr(level_1_dt, label = TRUE, label_round = 2)
```


```{r}
# test set performances
print(h2o.auc(h2o.performance(glm_model, newdata = data_test)))
print(h2o.auc(h2o.performance(gbm_model, newdata = data_test)))
print(h2o.auc(h2o.performance(deeplearning_model, newdata = data_test)))
```

```{r}
# for the ensemble model
print(h2o.auc(h2o.performance(ensemble_model, newdata = data_test)))
```

The baseline meta-learner is a glm model. You can try others:
```{r}
ensemble_model_gbm <- h2o.stackedEnsemble(
  X, y,
  training_frame = data_train,
  metalearner_algorithm = "gbm",
  base_models = list(glm_model, 
                     gbm_model,
                     deeplearning_model))
```

```{r}
print(h2o.auc(h2o.performance(ensemble_model_gbm, newdata = data_test)))
```

Meta-learning can also be built upon same-family, different
hyperparameter models via a grid of hyperparameters.

```{r}
learn_rate_opt <- c(0.1, 0.3)
max_depth_opt <- c(3, 5, 7)
hyper_params <- list(learn_rate = learn_rate_opt,
                     max_depth = max_depth_opt)

gbm_grid <- h2o.grid(
  x = X, y = y,
  training_frame = data_train,
  algorithm = "gbm",
  ntrees = 10,
  hyper_params = hyper_params,
  seed = 123,
  nfolds = 5, 
  keep_cross_validation_predictions = TRUE)
```

```{r}
ensemble_model_grid_gbm <- h2o.stackedEnsemble(
  X, y,
  training_frame = data_train,
  base_models = gbm_grid@model_ids)
```
```{r}
# individual test set performances
for (model_id in gbm_grid@model_ids) {
  model <- h2o.getModel(model_id)
  print(model_id)
  print(h2o.auc(h2o.performance(model, newdata = data_test)))
}
```

```{r}
print(h2o.auc(h2o.performance(ensemble_model_grid_gbm, newdata = data_test)))
```

## AutoML

Based on stacking H2O offers an out-of-the-box tool called "auto ML". It runs some
models that differ from each other and then creates a stacked version of them. Finally,
the best model is selectd.
```{r}
automl_model <- h2o.automl(X, y,
  training_frame = data_train,
  max_runtime_secs = 120)  # you can specify how long (approx.) you want the estimation to run
```

"Leaderboard" of the models: how well they perform based on cross-validation.
```{r}
lb <- automl_model@leaderboard
print(lb, n = nrow(lb))
```

```{r}
# the best performing model is stored in '@leader'
automl_model@leader
```

```{r}
print(h2o.auc(h2o.performance(automl_model@leader, newdata = data_test)))
```

For a more in-depth tutorial, see [here](https://github.com/h2oai/h2o-tutorials/tree/master/h2o-world-2017/automl).
